quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-08-26 paper-quollr.Rmd
Paper changes up to the linked plots section

@@

Non-linear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2019], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021], create wildly different representations depending on the selected method and hyper-parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

The following demonstration of the package's functionality assumes `quollr` has been loaded. We also want to load the built-in data sets `scurve` and `scurve_umap`.

`scurve` is a $7\text{-}D$ simulated dataset. It is constructed by simulating $1000$ observations from $\theta \sim U(-3\pi/2, 3\pi/2)$, $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$ (adding thickness to the S), $X_3 = \text{sign}(\theta) \times (\cos(\theta) - 1)$. The remaining variables $X_4$, $X_5$, $X_6$, $X_7$ are all uniform error, with small variance. `scurve_umap` is the UMAP $2\text{-}D$ embedding for `scurve` data with `n_neighbors` is $46$ and `min_dist` is $0.9$. Each data set contains a unique ID column that maps `scurve` and `scurve_umap`.

@@

Constructing the $2\text{-}D$ model primarily involves (i) scaling the NLDR data, (ii) binning the data, (iii) obtaining bin centroids, (iv) connecting centroids with line segments to indicate neighbors, and (v) Remove low-density hexagons.

@@

The algorithm starts by scaling the NLDR data to a standard range using the `gen_scaled_data()` function. This function standardizes the data so that the first embedding ranges from $0$ to $1$, while the second embedding scales from $0$ to the maximum value of the second embedding. The output includes the scaled NLDR data along with the original limits of the embeddings.

@@

The configurations of a hexagonal grid are determined by the number of bins and the bin width in each direction. The function `calc_bins_y()` is used for this purpose. This function accepts an object containing scaled NLDR data in the first and second columns, along with numeric vectors that represent the limits of the original NLDR data, the number of bins along the x-axis (`b1`), and the buffer amount as a proportion (`q`).

@@

Points are allocated to bins based on the nearest centroid. The hexagonal binning algorithm can be executed using the `hex_binning()` function, or its components can be run separately for added flexibility. The parameters used within `hex_binning()` include an object containing scaled NLDR data in the first and second columns, along with numeric vectors that represent the limits of the original NLDR data (`nldr_obj`), the number of bins along the x-axis (`b1`), and the buffer amount as a proportion of the data (`q`). The output is an object of the `hex_bin_obj` class, which contains the bin widths in each direction (`a1`, `a2`), the number of bins in each direction (`bins`), the coordinates of the hexagonal grid starting point (`start_point`), the details of bin centroids (`centroids`), the coordinates of bins (`hex_poly`), NLDR components with their corresponding hexagon IDs (`data_hb_id`), hex bins with their corresponding standardized counts (`std_cts`), the total number of bins (`tot_bins`), the number of non-empty bins (`non_bins`), and the points within each hexagon (`pts_bins`).

@@

Then, the centroids are computed iteratively. The x-coordinates for centroids in odd-numbered rows are initialized as a sequence spaced by the hexagon width. Even-numbered rows are staggered by half this width to achieve a hexagonal tiling effect. Vertical spacing (`a2`) is derived by $\sqrt{3}/2 \times a_1$.

@@

With these values, the function determines fixed offsets in the x and y directions for all six vertices relative to the centroid. These offsets form two vectors (`x_add_factor` and `y_add_factor`) corresponding to the six compass directions used to define the polygon shape: top, top-left, bottom-left, bottom, bottom-right, and top-right.

@@

The `compute_std_counts()` function calculates both the raw and standardized counts for each hexagon.

The function begins by grouping the data by `h` and counting the number of NLDR points falling within each bin. These raw counts are stored as `n_h`. To enable comparisons across bins with varying densities, the function then standardizes these counts by dividing each bin’s count by the maximum count across all bins. This yields a normalized metric, `w_h`, ranging from 0 to 1.

@@

The `find_low_dens_hex()` function identifies hexagons with low point densities, considering the densities of their neighboring bins as well. The `find_low_dens_hex()` function relies on a helper, `compute_mean_density_hex()`, which iterates over all hexagons and computes the average density across neighbors based on their `h` and a defined number of bins along the x-axis (`b1`). The hexagonal layout assumes a fixed grid structure, so neighbor IDs are computed by positional offsets.

@@

The final step involves lifting the fitted $2\text{-}D$ model into $p\text{-}D$ by computing the $p\text{-}D$ mean of data points within each hexagonal bin to represent bin centroids. This transformation is performed using the `avg_highd_data()` function, which takes $p\text{-}D$ data (`highd_data`) and embedding data with their corresponding hexagonal bin IDs as inputs (`scaled_nldr_hexid`).

@@

The `predict_emb()` function is used to predict $2\text{-}D$ embedding for a new $p\text{-}D$ data point using the fitted model. This function is useful to predict $2\text{-}D$ embedding irrespective of the NLDR technique.

In the prediction process, first, the nearest $p\text{-}D$ model point is identified for a given new $p\text{-}D$ data point by computing $p\text{-}D$ Euclidean distance. Then, the corresponding $2\text{-}D$ bin centroid mapping for the identified $p\text{-}D$ model point is determined. Finally, the coordinates of the identified $2\text{-}D$ bin centroid is used as the predicted NLDR embedding for the new $p\text{-}D$ data point.

@@

As a Goodness of fit statistics for the model, `glance()` is used to compute residuals and RMSE. These metrics are used to assess how well the fitted model will capture the underlying structure of the $p\text{-}D$ data.

@@

The function starts by renaming columns in the `model_highd` data frame to avoid conflicts. It then predicts the high-dimensional bin assignments for each point using the `predict_emb()` function, mapping each observation to its nearest $2\text{-}D$ bin (`pred_hb_id`). This prediction is joined with the mean high-dimensional coordinates of each bin from the model and with the original high-dimensional data.

@@

To evaluate how well the $p\text{-}D$ model captures the underlying structure of the high-dimensional data, we provide a visualization using the `show_langevitour()` function. This function renders a dynamic projection of both the high-dimensional data and the model using the `langevitour` package.

>>>>

::: {.hidden}
$$
\newcommand{\pD}{$p\text{-}D$}
\newcommand{\twoD}{$2\text{-}D$}
$$
:::

@@

Non-linear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2019], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021], can create hugely different representations depending on the selected method and hyper-parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

The following demonstration of the package's functionality assumes `quollr` has been loaded. The built-in data sets `scurve` and `scurve_umap` are also used throughout.

`scurve` is a $7\text{-}D$ simulated dataset. It is constructed by simulating $1000$ observations from

$$
\begin{aligned}
\theta &\sim U(-3\pi/2, 3\pi/2), \\
X_1 &= \sin(\theta), \\
X_2 &\sim U(0, 2), \text{adding thickness to the S}, \\
X_3 &= \text{sign}(\theta) \times (\cos(\theta) - 1).
\end{aligned}
$$

The remaining variables $X_4$, $X_5$, $X_6$, $X_7$ are all uniform error, with small variance. `scurve_umap` is the UMAP $2\text{-}D$ embedding for `scurve` data with `n_neighbors`$=46$ and `min_dist`$=0.9$. Each data set contains a unique ID column that maps the $7\text{-}D$ point in `scurve` to the corresponding $2\text{-}D$ `scurve_umap`.

@@

Constructing the $2\text{-}D$ model primarily involves (i) scaling the NLDR data, (ii) binning the data, (iii) obtaining bin centroids, (iv) connecting centroids with line segments to indicate neighbors, and (v) removing low-density hexagons.

@@

The algorithm starts by scaling the NLDR data to to the range $[0, 1] \times [0, x_2]$, where $x_2$ is the maximum value of the second embedding, using the `gen_scaled_data()` function. The output includes the scaled NLDR data (`scaled_nldr`) along with the original limits of the embeddings (`lim1`, `lim2`).

@@

The configurations of a hexagonal grid are determined by the number of bins and the bin width in each direction. The function `calc_bins_y()` is used for this purpose. This function accepts (1) an object containing three elements: the first is the scaled NLDR data with two columns, and the second and third are the limits of the original NLDR data; (2) the number of bins along the x-axis (`b1`), and (3) the buffer amount as a proportion (`q`).

@@

Points are allocated to bins based on the nearest centroid of the hexagonal bins. The hexagonal binning algorithm can be executed using the `hex_binning()` function, or its components can be run separately for added flexibility. The parameters used within `hex_binning()` include an object containing scaled NLDR data in the first and second columns, along with numeric vectors that represent the limits of the original NLDR data (`nldr_obj`), the number of bins along the x-axis (`b1`), and the buffer amount as a proportion of the data (`q`). The output is an object of the `hex_bin_obj` class, which contains the bin widths in each direction (`a1`, `a2`), the number of bins in each direction (`bins`), the coordinates of the hexagonal grid starting point (`start_point`), the details of bin centroids (`centroids`), the coordinates of bins (`hex_poly`), NLDR components with their corresponding hexagon IDs (`data_hb_id`), hex bins with their corresponding standardized counts (`std_cts`), the total number of bins (`tot_bins`), the number of non-empty bins (`non_bins`), and the points within each hexagon (`pts_bins`).

@@

Then, the centroids are computed iteratively. The x-coordinates for centroids in odd-numbered rows are initialized as a sequence spaced by the hexagon width. Even-numbered rows are staggered by half this width to achieve a hexagonal tiling effect. Vertical spacing (`a2`) is given by the height of a hexagon, $\sqrt{3}/2 \times a_1$.

@@

With these values, the function determines fixed offsets in the x and y directions for all six vertices relative to the centroid. These offsets form two vectors corresponding to the six compass directions used to define the polygon shape: top, top-left, bottom-left, bottom, bottom-right, and top-right.

@@

The `compute_std_counts()` function calculates both the raw and standardized counts of points inside each hexagon.

The function begins by grouping the data by hexagon ID (`h`) and counting the number of NLDR points falling within each bin. These raw counts are stored as `n_h`. To enable comparisons across bins with varying densities, the function then standardizes these counts by dividing each bin’s count by the maximum count across all bins. This yields a normalized metric, `w_h`, ranging from 0 to 1.

@@

The `find_low_dens_hex()` function relies on a helper, `compute_mean_density_hex()`, which iterates over all hexagons and computes the average density across neighbors based on their hexagon ID (`h`) and a defined number of bins along the x-axis (`b1`). The hexagonal layout assumes a fixed grid structure, so neighbor IDs are computed by positional offsets.

@@

The final step involves lifting the fitted \twoD model into \pD. This is done by modelling a point in \pD as the  \pD mean of data points in the \twoD centroid. This is performed using the `avg_highd_data()` function, which takes $p\text{-}D$ data (`highd_data`) and embedding data with their corresponding hexagonal bin IDs as inputs (`scaled_nldr_hexid`).

@@

The `predict_emb()` function is used to predict a point in a $2\text{-}D$ embedding for a new $p\text{-}D$ data point using the fitted model. This function is useful to predict $2\text{-}D$ embedding irrespective of the NLDR technique.

In the prediction process, first, the nearest $p\text{-}D$ model point is identified for the new $p\text{-}D$ data point by computing $p\text{-}D$ Euclidean distance. Then, the corresponding $2\text{-}D$ bin centroid mapping for the identified $p\text{-}D$ model point is determined. Finally, the coordinates of the identified $2\text{-}D$ bin centroid is used as the predicted NLDR embedding for the new $p\text{-}D$ data point.

@@

Residuals and root mean squared error (RMSE) are used as goodness of fit metrics for the model. These metrics can be output using the  `glance()` function.

@@

The function starts with the same process as is used in the `glance()` function to produce a predicted point in \pD for each point in the high-dimensional dataset.

@@

To visualise how well the $p\text{-}D$ model captures the underlying structure of the high-dimensional data, we provide a tour of the model in \pD using the `show_langevitour()` function. This function renders a dynamic projection of both the high-dimensional data and the model using the `langevitour` package.
